A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-23 latex/main.tex
Comparison with RIPPLES methods

\subsubsection*{Comparison with UShER+RIPPLES}
We began with the Viridian UShER tree downloaded from FigShare.
To simplify the comparison with sc2ts, we subset this 
file down to the 2,475,418 samples shared with the sc2ts ARG 
by running \texttt{matUtils extract}.
We then ran \texttt{ripples-fast}~\cite{Smith2023tracking}
on this file, using \texttt{-n1} so that all detected 
recombinants were reported (equivalent to sc2ts). We ran
ripples with the parsimony improvement parameter $p$
equal to 3 (the default) and 4.

The output of RIPPLES is two text files,

To compare these results with the sc2ts ARG, we first matched clades
by finding the most recent common ancestor (MRCA) of the samples
descending from each RIPPLES event. Overall, clade agreement was
high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
where descendants match exactly. Of these nodes, 68\% had no recombinant
ancestry. Of the remainder, 595 could be matched with
one of the 855 recombination events in the sc2ts ARG

>>>>

\subsubsection*{Comparison with UShER+RIPPLES}
We began with the Viridian UShER tree downloaded from FigShare.
To simplify the comparison with sc2ts, we subset this 
file down to the 2,475,418 samples shared with the sc2ts ARG 
by running \texttt{matUtils extract}.
We then ran \texttt{ripples-fast}~\cite{Smith2023tracking}
on this file, using \texttt{-n1} so that all detected 
recombinants were reported (equivalent to sc2ts). We ran
RIPPLES with the parsimony improvement parameter $p$
equal to 3 (the default) and 4.

The output of RIPPLES is two text files, ``recombination.tsv''
and ``descendants.tsv``. We use the ``descendants.tsv'' file,
which lists the descendant samples of all recombination events
keyed by the recombinant node ID, to match events with sc2ts.
Specifically, for each RIPPLES event identified a list of 
descendant samples, we find the MRCA of 
that set of samples in the first tree of the sc2ts ARG. 
The clades are then said to agree if the number of additional 
samples descending from the sc2ts node is $<2$
(Figure~\ref{fig:ripples_sc2ts_events}).

We ran 3SEQ on the RIPPLES recombinants by extracting the 
alignments for the child and parent sequences for each 
recombination event. We first obtained the parent
node IDs from the ``recombination.tsv'' file by choosing 
the most parsimonious event associated with each recombinant
node,  arbitrarily choosing among the possibilities in 
the case of ties. We then used this list of child and 
parent node IDs as 
input to \texttt{matUtils extract -v} to extract their
sequences in VCF format. Following this, 
we used vcflib~\cite{Garrison2022spectrum} to convert
to alignments in FASTA format, as recommended by
the UShER documentation (\texttt{vcf2fasta -f NC\_045512v2.fa}).
We then ran 3SEQ in ``full mode''
independently for each trio, in the same manner as described
when validating the sc2ts recombinants.

The UShER tree is not dated, and so to obtain dates for 
parent nodes we used sample metadata 
(Date\_Tree from the Viridian metadata). For the minority 
of parent sequences that are not samples, we used the oldest
sample descending from a given internal node as a proxy for 
its date. We used the BTE Python library 
(\url{https://jmcbroome.github.io/BTE/build/html/index.html})
to perform the required traversals on the UShER tree.

The full workflow is encoded in the Snakemake postprocessing 
pipeline.
See Document~\ref{sec:detailed_ripples} for detailed analysis
of the RIPPLES events, 
and the
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_ripples.ipynb}{analysis\_ripples.ipynb}
notebook for full methodological details of the analysis.
